Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-03-11 agujournaltemplate.tex
Discussion draft 40%

    % Phytoplankton is important
    Phytoplankton provides the foundation for marine food webs and plays a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them highly suitable indicators of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
    The structure of the community can be considered an integrator of environmental factors, as it is strongly determined by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}. In addition to such bottom-up effects, there are also forces of higher trophic levels, such as zooplankton, that structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}. Disentangling the effects driving changes in communities and biodiversity of phytoplankton is a difficult but crucial task to better understand our oceans.

    % Phytoplankton is under threat
    On a global scale, human activities are driving a decline in biodiversity \cite{tittensor_mid-term_2014}. Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}.  The drivers and effects of changes in biodiversity need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014} and is under greater threat of extinction, compared to temperate regions \cite{finnegan_paleontological_2015}.

@@

    % Describe CARIACO changes in ecosystem
    Since the inception of the time series in 1995, the ecosystem of the Cariaco basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors related to the impact of global warming. The phytoplankton community identified by microscopy showed a drastic reduction in Diatom abundance and a shift toward smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
    The trends in microscopy community data were confirmed and expanded by \citeA{pinckney_phytoplankton_2015}. They observed that the drastic reduction in diatoms counted using microscopy corresponded, but to a lesser degree, to diatom abundance calculated from the pigment composition measured by high-performance liquid chromatography (HPLC). This is most likely explained by a shift to diatom species with smaller cell sizes, which are more difficult to detect and identify using light microscopy \cite{lorenzoni_characterization_2015}. In addition, the pigment data indicate a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in the community composition. These changes are most likely a response to the reduced upwelling and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.
    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data past 2012 until the end of the CARIACO time series in 2017, which was not included in previous studies of the phytoplankton community, indicates a return to stronger upwelling intensity and similar productivity levels to the late 1990s \cite{muller-karger_scientific_2019}.

@@

    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the data set more homogeneous and robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008}. Furthermore, we calculate the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

@@

% Figure 1
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celcius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

% Figure 2
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008 and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall. 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 3
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. Upwelling related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celcius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

% Figure 4
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to identify if time lags of the predictor variables could better explain the data (see Appendix), and this resulted in choosing the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capacity for climate indices with time lags and upwelling-related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower. The \qty{21}{\celsius} Isotherm and silicate and phosphate concentrations scored lowest in predicting changes within the community. The way Gradient Forest predicts shifts in the community is by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data are biased towards larger cell sizes, and therefore most of the predicted genus cell counts are from diatoms. However, two Haptophytes and four Dinoflagellates, as well as the Cyanobacteria \textit{Synechococcus} do rank highly.
Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range. The black line indicates cumulative importance across the entire community and the colored dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies highly affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes. Nitrate concentrations between 6 and 8 \unit{\micro \mole} is where the community changed most, indicating these values as a threshold for the composition of the community. For wind speed, temperature, and salinity, changes occurred most at parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}). An interesting aspect is that, for both climate indices, haptophytes showed the strongest cumulative importance, but for diatoms, the nitrate concentration seemed to be a stronger predictor.

@@

% HERE I INTERPRET AND DISCUSS, whereas before I just described. Answer not the *what*, but the *how* and *why*.
Our main results show that...
1. we see Overlap in yearly mean anomalies between certain variables

- The Cariaco basin is a system, that...
    - Discuss cell count collapse, apparent shift, but not as strong! Limitations of microscopy.

This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes. In the Cariaco basin, long-term climatic oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
%need to rephrase below (from Pinckney)
During the early years of the time series (1996–2004), the Cariaco basin experienced a regular seasonal upwelling cycle driven by strong Trade Winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. This upwelling period represented \qty{\sim 70}{\%} of annual depth-integrated primary production with a phytoplankton community dominated by microphytoplankton (diatoms, dinoflagellates, and coccolithophorids) \cite{pinckney_phytoplankton_2015}.
Following 2004, the physical and biogeochemical regime had changed to one with significantly reduced Trade Winds during the upwelling period, resulting in higher sea surface temperatures and less pronounced mixing of the surface and deep layers \cite{taylor_ecosystem_2012}. The microscopy cell counts indicate a collapse of large phytoplankton and diatoms in particular, which is confirmed to a lesser degree by pigment data analysis. In general, a shift in the phytoplankton community to smaller cell sizes and higher abundances of coccolithophores, cryptophytes, and other phytoflagellates, typical for stratified water columns was observed \cite{pinckney_phytoplankton_2015}.

- Now mention how the variables overlap in anomalies in Figure 2, how that can be linked to the general changes already observed by other studies.

@@

3. Clustering shows return of community to previous cluster, confirmed by monthly NMDS, but EnvFit is interesting, as no variable clearly goes along gradient separating the two cluster (NMDS2)

3.1 Turnover is different between two clusters, also corresponding to evennness between clusters

@@

@@

    - AMO (ref to 2019 Review)
    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al. 2015). These climatic changes had a profound impact throughout the Atlantic (Alexander et al. 2014, Nye et al. 2014) and in the Cariaco basin \cite{taylor_ecosystem_2012,pinckney_phytoplankton_2015}.
    - ENSO teleconnection (ref to 2019 Review)
    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).

5. Compare to what happened at other long term time series in the ocean...

%"Taylor et al. 2012 observed that the decline in cell abundances was disproportionate compared with the smaller decrease in total chlorophyll a, which implies that phytoplankton of smaller size were not detected in the surveys.. This was supported by high-performance liquid chromatography analyses, which indicated a reduction in diatom concentrations and an increase in phytoplankton diversity. A return to a phytoplankton community structure similar to that observed in the late 1990s was observed in 2014 and 2015 (Figure 5a), as upwelling intensity increased to levels similar to those seen before 2003." from 2019 Review Legacy of CARIACO

>>>>

    % Phytoplankton is important
    Phytoplankton provides the foundation for marine food webs and plays a crucial role in biogeochemical processes, including oxygen production and carbon export to the deep ocean \cite{falkowski_biogeochemical_1998}.
    Phytoplankton consists of a diverse set of organisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them highly suitable indicators of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.
    The structure of the community can be considered an integrator of environmental factors, as it is strongly determined by temperature, nutrient availability, and light availability \cite{mouw_phytoplankton_2016}. In addition to such bottom-up effects, there are also forces of higher trophic levels, such as zooplankton, that structure the community and affect biodiversity through grazing pressure \cite{banas_adding_2011}. Disentangling the effects driving changes in community composition and biodiversity of phytoplankton is a difficult but crucial task to better understand our oceans.

    % Phytoplankton is under threat
    On a global scale, human activities are driving a decline in biodiversity \cite{tittensor_mid-term_2014}. Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}.  The drivers and effects of changes in biodiversity need to be further investigated, particularly in the tropics \cite{clarke_does_2017}, where community diversity is often highest \cite{brown_why_2014, righetti_global_2019} and is under greater threat of extinction, compared to temperate regions \cite{finnegan_paleontological_2015}.

@@

    % Describe CARIACO changes in ecosystem
    Since the inception of the time series in 1995, the ecosystem of the Cariaco basin has undergone marked changes. \citeA{taylor_ecosystem_2012} analyzed time series data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling and phytoplankton bloom intensities, which the authors related to the impact of global warming. The phytoplankton community identified by microscopy showed a drastic reduction in Diatom abundance and a shift toward smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass.
    The trends in microscopy community data were partially confirmed and expanded by \citeA{pinckney_phytoplankton_2015}. They observed that the drastic reduction in diatoms counted using microscopy corresponded, but to a lesser degree, to diatom abundance calculated from the pigment composition measured by high-performance liquid chromatography (HPLC). This is most likely explained by a shift to diatom species with smaller cell sizes, which are more difficult to detect and identify using light microscopy \cite{lorenzoni_characterization_2015}. In addition, the pigment data indicate a deepening of the chlorophyll maximum, an increase in diversity, and a reduction in seasonal variability in the community composition. These changes are most likely a response to the reduced upwelling and a deepening of the euphotic zone associated with a decrease in the strength of seasonal trade winds. Interestingly, changes in the phytoplankton community did not result in a significant change in the export of particulate organic carbon (POC) to depth, as measured by sediment traps deployed during the time series \cite{pinckney_phytoplankton_2015}.
    %%% Maybe add a point about the Chlorophyll being low and high, HPLC vs Fluorometric?
    The shift in the community, due to the warming trend and the reduction in upwelling intensities, does not appear to be a permanent regime shift. Data past 2012 until the end of the CARIACO time series in 2017, which was not included in previous studies of the phytoplankton community, indicates a return to stronger upwelling intensity and similar productivity levels to the late 1990s \cite{muller-karger_scientific_2019}.

@@

    As the main indicator of diversity, we used genus richness, thereby reducing taxonomic resolution to make the data set more homogeneous and robust to discrepancies in sampling effort \cite{ptacnik_diversity_2008, de2020higher}. Furthermore, we calculate the Shannon index $H' = -\sum_i p_i \log_{e} p_i$ and Pielou\textquotesingle s evenness $J' = -\sum_i p_i ln( p_i )/ln(H')$, where $p_i$ represents the genera observed for each interpolated and integrated monthly sampling.

@@

% Figure 2
The normalized yearly means for all environmental variables considered, as well as the logarithmic transformation of chlorophyll \textit{a} and phytoplankton cell counts per functional group, are shown in Figure \ref{fig:zscore}. The patterns emerging from the -u 10m wind component, the temperature and the directly related \qty{21}{\celcius} isotherm clearly show the years of positive anomalies of wind (blue), upwelling (blue, i.e., shallower depth of the isotherm) and negative for temperature (red) within the top 100 m. These patterns are reflected in the mode of the AMO, where a positive anomaly seems to correlate with weak upwelling regimes. The yearly mean nutrient concentrations have a maximum positive anomaly at the start of the time series, but show differing inter-annual oscillations. Annual mean anomalies of nitrate and phosphate seem to follow upwelling anomalies but remain relatively low between 2002 and 2014, with some more variability in mean phosphate concentration. Silicate does not follow a clear pattern, but shows an elevated mean concentration toward the later part of the time series (between 2008 and 2013), which could be driven by increased surface runoff \cite{lorenzoni_characterization_2015} and reduced depletion via diminished Diatom abundances.

% Figure 3
To further investigate the depth distribution and dynamics of chlorophyll \textit{a} and community diversity, Figure \ref{fig:divts} shows the depth distributions over time, as well as between seasons. The bulk of the chlorophyll biomass ($\sim$ 80\%) was measured in the upper 50 meters of the water column, with approximately 15\% measured between 50 and 75 meters. Chlorophyll shows marked seasonality, with the highest biomass in the upper water column during the winter months (December–February, see Figure \ref{fig:divts} $b$). The yearly mean chlorophyll biomass shows the highest variability within the upper 25 meters. In some years, particularly during 2000, 2005, 2007, 2008 and 2013, its concentration in this layer was lower than in the 25–50 meter depth range (see Figure \ref{fig:divts} $a$). In general, yearly mean anomalies within the upper 75 meters appear to be elevated during the period 1996 to 2004 and only recover in 2012. The year 2013 is an outlier, particularly for the upper 25 meters, as the missing monthly samplings from the winter month (January and February) skew the value. Nevertheless, the monthly measurements for the remainder of the year 2013 show a markedly lower biomass (65\% of the mean over the entire time series for these months), which recovers in 2014. Chlorophyll biomass below 50 meters depth shows a relatively stable distribution throughout the year and across the time series. In Fall, during the rainy season, mean chlorophyll biomass in the upper 25 meters is markedly reduced, while the 50–75 meters depth interval sees a slight increase, showing a shift in the chlorophyll maximum to lower depth under weak upwelling conditions.
The calculated genus diversity from the microscopy cell counts generally follows the dynamics of chlorophyll biomass in seasonality, with maximum values observed during the upwelling season in winter. Genus richness is consistently highest in the top \qty{25}{meters} depth interval, which also shows the most variability over the time series. Across the time series, we have maximum richness at the start of the time series (1996-1997), then a reduction in diversity across all depths, which only recovers past 2011. The collapse in diversity following the start of the time series does not follow the signal shown by chlorophyll \textit{a} biomass, which remains relatively stable across the top \qty{100}{meters} until 2004 (see Figure \ref{fig:divts $a$}). Scaling this genus diversity by the cell count abundances, i.e. the Shannon index, shows a similar but less pronounced dynamic. Diversity remains high throughout 2000 and recovers more quickly, but shows marked reductions in annual means, particularly in the upper 25 meters in 2006, 2009 and 2016. The seasonal signal is almost lost (see Figure \ref{fig:divts} $f$), yet there is a consistent pattern of higher Shannon diversity at lower depths for all seasons except for Fall. 
Community evenness as signified by the Pielou index shows no seasonal signal and a very large variance within seasons (see Figure \ref{fig:divts} $h$). The inter-annual mean dynamics of evenness are remarkably stable and similar across all depth intervals, clustered around an intermediate 0.5 value. There are marked reductions in all depths towards 2000 and for all depths above 75 meters in 2006, 2009 and 2016. Throughout 2003 to 2013 evenness is almost consistently highest in the deepest interval below 75 meters, where the community seems to have been more evenly distributed.

% Figure 4
Taking the microscopy community data, interpolated within the top 100 meters, we wanted to find if there was an apparent structure within the time series that could be related to the physical environment. For this, we used different clustering methods, both on the annual aggregate of cell counts from microscopy and on the individual monthly measurements. Figure \ref{fig:clustering} $a$ shows the results of clustering the annual aggregate communities with a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster including the years 1996-2004 and 2014-2016 and the second cluster of the intermediate years from 2004-2013. 
Inter-annual species turnover is shown in Figure \ref{fig:clustering} $b$, where there is an increasing trend in turnover until 2004, where turnover is reduced for the period of cluster 2, but in 2014 with the return to cluster 1, the turnover is higher again. This pattern is also reflected in the clustering in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. 
To check whether annual aggregation of the community data smoothed out strong seasonal signals, we performed an NMDS cluster analysis using the monthly cell counts (see Figure \ref{fig:clustering} $c$). This NMDS clustering is overlaid with environmental data and climate indices to help explain some of the underlying trends. The first axis separates the two clusters relatively cleanly, where cluster 2 is trending towards positive AMO anomalies and cluster 2 trends towards positive anomalies in salinity and the MEI v.2 index. The second NMDS axis appears to follow the trends in upwelling intensity as described by the opposing trend arrows for nutrient concentrations and temperature, as well as wind speed. Overall, NMDS clustering seems to confirm the pattern found by the annual aggregate clustering. 
Figure \ref{fig:clustering} $d$ to $o$ shows density distributions for environmental variables, climate and diversity indices between the two clusters. Upwelling related variables show distinct patterns between the two clusters. For example, cluster 1 experiences more frequent high wind events, a shallower \qty{21}{\celcius} Isotherm, lower temperature, and higher salinity and nutrient concentrations, in particular for nitrate. The AMO index shows a striking separation between the two clusters, where cluster 1 shows more negative anomalies, while cluster 2 has a high occurrence of positive anomalies. Although the mean diversity for all the indices is relatively similar, with a slightly higher mean in the Shannon and Pielou index for cluster 2, the density distribution shows marked differences. For cluster 1 there is a higher variance in genus richness, while cluster 2 shows higher variance and a bi-modal distribution for Shannon and Pielou indices.

% Figure 5
To identify the variables that best predict the shifts observed within the phytoplankton community, we performed a Gradient Forest analysis on the microscopy cell counts interpolated over the top 100 meters and grouped by identified genus. Preliminary model runs were used to identify if time lags of the predictor variables could better explain the data (see Appendix), and this resulted in choosing the AMO with a lag of 2 months and MEI v.2 with a lag of 4 months. In Figure \ref{fig:GF} $a$, we show the predictor variables ranked by their weighted importance, with the highest predictive capacity for climate indices with time lags and upwelling-related variables such as nitrate in the upper 100 meters and wind speed scoring slightly lower. The \qty{21}{\celsius} Isotherm and silicate and phosphate concentrations scored lowest in predicting changes within the community. The way Gradient Forest predicts shifts in the community is by building a random forest model for each observed genus across the time series. The predictive power of the models for each genus is shown in Figure \ref{fig:GF} $b$. The microscopy data are biased towards larger cell sizes, and therefore most of the predicted genus cell counts are from diatoms. However, two Haptophytes and four Dinoflagellates, as well as the Cyanobacteria \textit{Synechococcus} do rank highly.
Figure \ref{fig:GF} $c$ shows the cumulative importance of the predictor variables across the parameter range. The black line indicates cumulative importance across the entire community and the colored dashed lines indicate importance per phytoplankton group. For the AMO index, we see a clear pattern in which negative anomalies highly affect the community. This contrasts with a higher importance for positive anomalies in the MEI v.2 index, particularly for haptophytes. Nitrate concentrations between 6 and 8 \unit{\micro \mole} is where the community changed most, indicating these values as a threshold for the composition of the community. For wind speed, temperature, and salinity, changes occurred most at parameter ranges related to high upwelling, i.e., high wind (between -4 and -6 \unit{m.s^{-1}}), low temperature (below \qty{23}{\celsius}) and high salinity (above \qty{36.8}{PSU}). An interesting aspect is that, for both climate indices, haptophytes showed the strongest cumulative importance, but for diatoms, the nitrate concentration seemed to be a stronger predictor.

@@

% HERE I INTERPRET AND DISCUSS, whereas before I just described. Answer not the *what*, but the *how* and *why*.
This present study demonstrates the highly dynamic nature of a tropical coastal marine ecosystem driven by local environmental variability, which, in turn, is affected by large-scale climatic changes. In the Cariaco basin, long-term climate oscillations have affected the wind-driven upwelling regime, which had confounding effects on the composition and biodiversity of the phytoplankton community.
During the initial phase of the time series (1996–2004), the Cariaco basin underwent a consistent seasonal upwelling cycle, driven by strong trade winds during winter and spring (December – April) \cite{mullerkarger_annual_2001, astor_seasonal_2003}. In this period, approximately \qty{\sim 70}{\%} of the annual depth-integrated primary production was observed. During upwelling, the phytoplankton community was dominated by microphytoplankton, consisting primarily of diatoms, dinoflagellates, and coccolithophorids \cite{pinckney_phytoplankton_2015}.
Following 2004,  there was a transition in the physical and biogeochemical regime characterized by a significant reduction in Trade Winds during the upwelling period, which led to elevated sea surface temperatures and decreased mixing between the surface and deep ocean layers \cite{taylor_ecosystem_2012}. Microscopy cell counts indicate the collapse of large phytoplankton, particularly diatoms, which is confirmed to a lesser extent by pigment data analysis. In general, a shift was observed in the phytoplankton community toward smaller cell sizes, along with increased abundances of coccolithophores, cryptophytes, and other phytoflagellates, which are typical of stratified water columns \cite{pinckney_phytoplankton_2015}.

We found that including the time series data up to its end in 2017 in our analysis clearly shows that the observed shifts were reversible. With a return to higher upwelling conditions in 2014, we also see a corresponding increase in biomass (see Figure \ref{fig:divts} $a$) and a return to a community more similar to the period between 1996 and 2004 (see Figure \ref{fig:clustering} $a$ and $c$). 
This shift to a more productive phytoplankton community after 2014 also coincided with a recovery in sardine capture observed in 2016 and 2017 in a location south-east of the Isla de Magarita \cite{gomez_gaspar_variacion_2025}.

When comparing the community between the two clusters...
- In microscopy counts we see a stark reduction in cell counts, which corresponds to a reduced concentration of chlorophyll \textit{a} biomass particularly in the top \qty{25}{meters} of the water column. When including the HPLC pigment data, these patterns are less clear, as the total chlorophyll measurement (check Pinckneys discssion on this again) actually shows an increase...
- The community shows a stark reduction in diatom cell counts, but also other functional groups --> shift towards smaller cell sizes, not observable in light microscopy
- Discuss why there is a shift to smaller cell sizes at low upwelling -> increased nutrient efficiency, cite some papers here.
- Discuss the dominant (most observed) genera per Cluster (TODO with DATA Analysis) and how that relates to the conditions
- Discuss how similar Part 1 and 2 of Cluster 1 are.. why are they clustered together?
- Discuss why we do not observe a clear distinction in Diversity betweent he two clusters, but mention the species turnover difference and the community evenness increase.

When comparing the environmental conditions between the two clusters...

%"Taylor et al. 2012 observed that the decline in cell abundances was disproportionate compared with the smaller decrease in total chlorophyll a, which implies that phytoplankton of smaller size were not detected in the surveys.. This was supported by high-performance liquid chromatography analyses, which indicated a reduction in diatom concentrations and an increase in phytoplankton diversity." from 2019 Review

%"A return to a phytoplankton community structure similar to that observed in the late 1990s was observed in 2014 and 2015 (Figure 5a), as upwelling intensity increased to levels similar to those seen before 2003." from 2019 Review Legacy of CARIACO

These changes are well reflected in the yearly mean anomalies shown in figure \ref{fig:zscore}. After 2004 we see higher SST, lower wind speed, reduced fluorometric chlorophyll \textit{a}, a collapse in functional group cell counts, all matching a shift in the AMO toward positive mean anomalies until 2014.
However, the shift in yearly mean genus richness does not follow the abundances in the cell counts. Instead, we see a marked reduction following 1998 up until 2010, where richness recovers four years before the community seems to return to previous levels in 2014. The same trend is apparent, although to a less pronounced degree, in the other diversity metrics. Generally, mean anomalies in Shannon diversity and Pielou's evenness correlate with genus richness and show a reduced yearly mean between 2000 and 2006, during which time the cell counts registered an increase in nanoflagellates and a pronounced peak in cyanobacteria abundance. 
The period between 2001 and 2006 lacks coverage in HPLC data, so we have to rely on microscopy data to deduce changes in the phytoplankton community. This limits the size range of the observable phytoplankton and thus the measurable diversity. We could deduce that we see a marked reduction in the diversity of microphytoplankton ahead of the collapse in abundance, and similarly an increase in microphytoplankton diversity a few years ahead of the return in abundance to previous levels.

- major thing is that previous paper only looked at microscopy to 55m (Mutshinda to 75m and Pinckney), but now we look at full 100 m!

- Seasonality and Depth Distribution of biomass and diversity.
The diversity of phytoplankton peaks in the surface ocean, similar to the patterns observed using genetic methods \cite{yeh_contrasting_2022}.

@@

3. Clustering shows the return of the community to the previous cluster, confirmed by monthly NMDS, but EnvFit is interesting, as no variable clearly follows the gradient separating the two clusters (NMDS2).

3.1 Turnover is different between two clusters, also corresponding to evenness between clusters

@@

    - We group species and genus counts, to genus level, and use genus counts as variables the model predicts. This assumes that there are not very specific niches etc. within genus, but that they are homogenous --> caveat!

@@

    - AMO (ref to 2019 Review)
    As the CARIACO time series was implemented in 1995, we now know that the Atlantic Multidecal Oscillation (AMO) was shifting to a positive phase for the next 20 years, which resulted in the largest oceanographic anomalies of the last century (McCarthy et al. 2015). These climatic changes had a profound impact throughout the Atlantic (Alexander et al. 2014, Nye et al. 2014) and in the Cariaco basin \cite{taylor_ecosystem_2012, pinckney_phytoplankton_2015}.
    - ENSO teleconnection (ref to 2019 Review)
    The Caribbean experiences the most marked of the El Ni  ̃no–Southern Oscillation teleconnections in the Atlantic Ocean, with warmer sea surface temperatures and trade winds blowing more to the north-northwest during the El Ni  ̃no–Southern Oscillation (Enfield & Mayer 1997).

5. Compare to what happened at other long-term time series in the ocean...
